---
output:
  pdf_document: default
  html_document: default
---

# Perform Enrichment Analysis

# Loading required libraries
```{r message=FALSE}

library(clusterProfiler)
library(org.Dm.eg.db)
library(pathview)
library(biomaRt)
library(ReactomePA)
library(ggplot2)
library(RColorBrewer)
library(ggsci)
```

# Read significant genes into csv file
```{r}
sig_genes <- function(file_path){
  genes <- read.csv(file_path, row.names = 1, stringsAsFactors = FALSE)
  final_genes <- na.omit(genes)
  return(final_genes)
  
}
significant_genes <- sig_genes("../Enrichment analysis in R/Data/Significant genes.csv")
head(significant_genes)
```

# Convert FlyBase IDs (FBgn) to ENTREZ IDs
```{r}
entrez_ids <- function(){
# copy the rownames of significant_genes and store it in flybase_ids
flybase_ids <- rownames(significant_genes)
# convert FLYBASE ids into ENTREZID
entrez_ids <- mapIds(org.Dm.eg.db, 
                     keys = flybase_ids, 
                     column = "ENTREZID", 
                     keytype = "FLYBASE", 
                     multiVals = "first")
# create column named ENTREZID in significant_genes that contain ENTREZID 
# of sig genes
significant_genes$ENTREZID <- entrez_ids
# remove ENTREZID that contain NA 
sign_genes <- significant_genes[!is.na(significant_genes$ENTREZID), ]
return(sign_genes)
}

sign_genes <- entrez_ids()
head(sign_genes)
```


# Up-regulated significant genes
```{r}
up_sig_genes <- function(){
  upregulated_genes <- sign_genes[ sign_genes$padj<0.01 & sign_genes$log2FoldChange>0 , ]
  return(upregulated_genes)
}

upregulated_sig_genes <- up_sig_genes()
upregulated_sig_genes
```


# ENTREZIDs of up-regulated significant genes
```{r}
up_entrez_ids <- upregulated_sig_genes$ENTREZID
up_entrez_ids
```


# Down-regulated significant genes
```{r}
down_sig_genes <- function(){
  downregulated_genes <- sign_genes[ sign_genes$padj<0.01 & sign_genes$log2FoldChange<0 , ]
  return(downregulated_genes)
}
downregulated_sig_genes <- down_sig_genes()
downregulated_sig_genes
```

# ENTREZIDs of Down-regulated significant genes
```{r}
down_entrez_ids <- downregulated_sig_genes$ENTREZID
down_entrez_ids

```


# Gene Ontology

## Group up-regulated significant genes that have similar BP GO terms 

```{r}

go_up <- function(){
    go <- groupGO( gene = up_entrez_ids,
                   OrgDb = org.Dm.eg.db,
                   ont = "BP",  # Biological Process
                   readable = TRUE)
    return(go)
  
}
go_terms_up <- go_up()
head(go_terms_up)

```
## Convert GO terms of up-regulated significant genes to DataFrame

```{r}
dataframe_go_up <- function(){
df_go_terms_up <- as.data.frame(go_terms_up)
return(df_go_terms_up)
}
df_go_group_up <- dataframe_go_up()
```

```{r}
df_go_group_up_top7 <- head(df_go_group_up,7)
```


## Barplot of up-regulated BP GO terms 
```{r}
bar_plot_up <- function(){
p <- ggplot(df_go_group_up_top7, aes(x = reorder(Description, - Count),
                                     y = Count, fill = Description)) +
                                     geom_bar(stat = "identity") +
                                     ggtitle("BP of up-regulated GO terms") +
                                     coord_flip() +
                                     theme_bw() +
                                     scale_fill_jama()+
                                     theme(plot.title = element_text(size = 12,
                                     face = "bold", hjust = 0.5))+
                                     xlab("Description")
jpeg("../Enrichment analysis in R/outputs/BPgroup_up_barplot.jpeg")
print(p)
dev.off()
}
bar_plot_up()
```

## Group down-regulated significant genes that have similar MF GO terms 
```{r}
go_down <- function(){
  go <- groupGO(gene = down_entrez_ids,
                OrgDb = org.Dm.eg.db,
                ont = "MF",  # Molecular function
                readable = TRUE)
  return(go)
}
go_terms_down <- go_down()
head(go_terms_down)
```
## Convert GO terms of down-regulated significant genes to DataFrame
```{r}
dataframe_go_down <- function(){
df_go_terms_down <- as.data.frame(go_terms_down)
return(df_go_terms_down)
}
df_go_group_down <- dataframe_go_down()
```

```{r}
df_go_group_down_top7 <- head(df_go_group_down,7)
```

## Barplot of down-regulated MF GO terms 
```{r}
bar_plot_down <- function(){
p <- ggplot(df_go_group_down_top7, aes(x = reorder(Description, - Count),
                                       y = Count, fill = Description)) +
                                       geom_bar(stat = "identity") +
                                       ggtitle("MF of down-regulated GO terms")+
                                       coord_flip() +
                                       theme_bw() +
                                       scale_fill_jama()+
                                       theme(plot.title = element_text(size = 12,
                                       face = "bold", hjust = 0.5))+
                                       xlab("Description")
jpeg("../Enrichment analysis in R/outputs/MFgroup_down_barplot.jpeg")
print(p)
dev.off()
}
bar_plot_down()
```


# Over-representation analysis

## Go enrichment analysis

## Enriched GO terms among up-regulated significant genes
```{r}
enrich_go_up <- function(){
        ego_up <- enrichGO( gene          = up_entrez_ids,
                            OrgDb         = org.Dm.eg.db,
                            keyType       = "ENTREZID",
                            ont           = "CC",
                            pvalueCutoff  = 0.05,
                            pAdjustMethod = "BH",   
                            readable      = TRUE)
        return(ego_up)
}
enrichment_go_up <- enrich_go_up()
head(enrichment_go_up)
```


## Dataframe of enriched GO terms for up-regulated significant genes
```{r}
df_go_terms_up <- function(){
  df_go_term_up <- as.data.frame(enrichment_go_up)
  return(df_go_term_up)
}
df_go_up <- df_go_terms_up()
```

```{r}
df_go_up_top7 <- head(df_go_up, 7)
```

## BarPlot of up-regulated CC enriched GO terms 
```{r}
bar_plot_enriched_up <- function(){
jama_colors <- pal_jama("default")(7)
p <- ggplot(df_go_up_top7, aes(x = reorder(Description, -Count),
                               y = Count, fill = p.adjust)) +
                               geom_bar(stat = "identity") +
                               coord_flip() +
                               scale_fill_gradientn(name = "p.adjust", 
                                                    colors = jama_colors) +
          labs(title = "Bar plot of CC of Enriched Up-regulated GO terms",
               x = "GO Term",
               y = "Gene Count") +
                              theme_bw() +
          theme(plot.title = element_text(size = 10,face = "bold", hjust = 0.5))
jpeg("../Enrichment analysis in R/outputs/CC_enriched_up_barplot.jpeg")
print(p)
dev.off()
}
bar_plot_enriched_up()
```


## Dotplot of up-regulated CC enriched GO terms
```{r}
dot_plot_enriched_up <- function(){
jama_colors <- pal_jama("default")(7)
p <- ggplot(df_go_up_top7, aes(x = reorder(Description, -Count),
                               y = Count, size = Count, color = p.adjust)) +
                               geom_point(alpha = 0.6) +
                               coord_flip() +
                               scale_size_continuous(range = c(3, 8), 
                                                     name = "Gene Count") +
                               scale_color_gradientn(name = "p.adjust", 
                                                     colors = jama_colors) +
                labs(title = "Dot plot of CC of enriched up-regulated genes",
                     x = "GO Term",
                     y = "Gene Count") +
                               theme_bw() +
                               theme(plot.title = element_text(size = 10, 
                                                               face = "bold", 
                                                               hjust = 0.5))
jpeg("../Enrichment analysis in R/outputs/CC_enriched_up_dotplot.jpeg")
print(p)
dev.off()
}
dot_plot_enriched_up()
```

## Network plot for enriched GO terms among up-regulated genes

```{r}
jpeg("../Enrichment analysis in R/outputs/Network_plot_up.jpeg")
cnet_plot_up <- cnetplot(enrichment_go_up, showCategory = 3, vertex.label.cex = 1.2)
print(cnet_plot_up)
dev.off()
```

## GO graph for enriched GO terms among up-regulated significant genes
```{r}
jpeg("../Enrichment analysis in R/outputs/GO_graph_up.jpeg")
go_graph_up <- plotGOgraph(enrichment_go_up)
print(go_graph_up)
dev.off()
```


## Enriched GO terms for down-regulated significant genes
```{r}
enrich_go_down <- function(){
        ego_down <- enrichGO( gene          = down_entrez_ids,
                            OrgDb         = org.Dm.eg.db,
                            keyType       = "ENTREZID",
                            ont           = "MF",
                            pvalueCutoff  = 0.05,
                            pAdjustMethod = "BH",   
                            readable      = TRUE)
        return(ego_down)

}

enrichment_go_down <- enrich_go_down()
head(enrichment_go_down)
```

## Dataframe of enriched GO terms among down-regulated significant genes
```{r}
df_go_terms_down <- function(){
df_go_term_down <- as.data.frame(enrichment_go_down)
return(df_go_term_down)
}
df_go_down <- df_go_terms_down()
```

## BarPlot of down-regulated MF enriched GO terms 
```{r}
bar_plot_enriched_down<- function(){
jama_colors <- pal_jama("default")(7)
p <- ggplot(df_go_down, aes(x = reorder(Description, - Count) ,
                            y = Count, fill = p.adjust)) +
                            geom_bar(stat = "identity") +
                            coord_flip() +
                            scale_fill_gradientn(name = "p.adjust", 
                                                 colors = jama_colors) +
            labs(title = "Bar plot of MF of Enriched down-regulated Genes",
                            x = "GO Term",
                            y = "Gene Count") +
                            theme_bw() +
                            theme(plot.title = element_text(size = 10, 
                                                            face = "bold", 
                                                            hjust = 0.5))
jpeg("../Enrichment analysis in R/outputs/MF_enriched_down_barplot.jpeg",
width = 700, height = 800)
print(p)
dev.off()
}
bar_plot_enriched_down()
```


## Dotplot of down-regulated MF enriched GO terms
```{r}
dot_plot_enriched_down<- function(){
jama_colors <- pal_jama("default")(7)
p <- ggplot(df_go_down, aes(x = reorder(Description, -Count),
                            y = Count, size = Count,
                            color = p.adjust)) +
                            geom_point(alpha = 0.6) +
                            coord_flip() +
                            scale_size_continuous(range = c(3, 8), 
                                                  name = "Gene Count") +
                            scale_color_gradientn(name = "p.adjust", 
                                                  colors = jama_colors) +
             labs(title = "Dot plot of MF of enriched down-regulated genes",
                            x = "GO Term",
                            y = "Gene Count") +
                            theme_bw() +
                            theme(plot.title = element_text(size = 10, 
                                                            face = "bold", 
                                                            hjust = 0.5))
jpeg("../Enrichment analysis in R/outputs/MF_enriched_down_dotplot.jpeg",
width = 700, height = 800)
print(p)
dev.off()
}
dot_plot_enriched_down()
```

## Network plot for enriched GO terms among down-regulated genes
```{r}
jpeg("../Enrichment analysis in R/outputs/Network_plot_down.jpeg")
cnet_plot_down <- cnetplot(enrichment_go_down, showCategory = 7, vertex.label.cex = 1.2)
print(cnet_plot_down)
dev.off()
```

## GO graph for enriched GO terms among down-regulated significant genes
```{r}
jpeg("../Enrichment analysis in R/outputs/GO_graph_down.jpeg")
go_graph_down <- plotGOgraph(enrichment_go_down)
print(go_graph_down)
dev.off()
```

## Pathway enrichment analysis

## Extract sig gene names from ensembl database 
```{r}

ensembl_gene_names <- function() {
  # Connect to the Ensembl database
  ensembl <- useMart("ensembl", dataset = "dmelanogaster_gene_ensembl")
  # list of Drosophila melanogaster FlyBase gene IDs
  flybase_gene_ids <- rownames(sign_genes)
  # Convert FlyBase gene IDs to external gene identifiers
  gene_info <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"), filters = "flybase_gene_id", values = flybase_gene_ids, mart = ensembl)
  return(gene_info)
}
gene_info_result <- ensembl_gene_names()
head(gene_info_result)
```

## Up-regulated Dmel_XXXX ids 
```{r}
dmel_up <- function(){
  filtered_up_sig_genes <- gene_info_result[gene_info_result$ensembl_gene_id %in%   rownames(upregulated_sig_genes), ]
  dmel_up_identifiers <- paste("Dmel", filtered_up_sig_genes$external_gene_name, 
                               sep = "_")
  return(dmel_up_identifiers)
}
dmel_up_genes <- dmel_up()
dmel_up_genes
```

## KEGG pathway enrichment analysis among up-regulated significant genes
```{r}
kegg_enrichment_up <- function(){
kegg_up <- enrichKEGG(gene = dmel_up_genes,
                        organism = "dme", # Drosophila melanogaster
                        pvalueCutoff = 0.05)
  return(kegg_up)
}
kegg_enrich_up <- kegg_enrichment_up()
head(kegg_enrich_up)
```

## DataFrame of KEGG enriched among up-regulated significant genes
```{r}
dataframe_kegg_up <- function(){
  df_kegg_up_genes <- as.data.frame(kegg_enrich_up)
  return(df_kegg_up_genes)
}
df_kegg_up <- dataframe_kegg_up()
head(df_kegg_up)
```

## Barplot for KEGG enriched up-regulated significant gene 
```{r}
bar_plot_kegg_enriched_up<- function(){
jama_colors <- pal_jama("default")(7)
p <- ggplot(df_kegg_up, aes(x = reorder(Description, - Count),
                            y = Count, fill = p.adjust)) +
                            geom_bar(stat = "identity", width = 0.8) +
                            coord_flip() +
                            scale_fill_gradientn(name = "p.adjust", 
                                                 colors = jama_colors) +
      labs(title = "Bar plot of Enriched KEGG pathway among up-regulated Genes",
           x = "Enriched pathway",
           y = "Gene Count") +
                            theme_bw() +
                            theme(plot.title = element_text(size = 10, 
                                                            face = "bold", 
                                                            hjust = 0.5))
jpeg("../Enrichment analysis in R/outputs/kegg_enriched_up_barplot.jpeg",
width = 700, height = 800)
print(p)
dev.off()
}
bar_plot_kegg_enriched_up()
```


## Dotplot for Enriched KEGG pathway among up-regulated significant gene
```{r}
dot_plot_kegg_enriched_up<- function(){
jama_colors <- pal_jama("default")(7)
p <- ggplot(df_kegg_up, aes(x = reorder(Description, -Count),
                            y = Count, size = Count,
                            color = p.adjust))+
                            geom_point(alpha = 0.6) +
                            coord_flip() +
                            scale_size_continuous(range = c(3, 8), 
                                                  name = "Gene Count") +
                            scale_color_gradientn(name = "p.adjust", 
                                                  colors = jama_colors) +
      labs(title = "Dot plot of Enriched KEGG pathway among up-regulated genes",
                            x = "GO Term",
                            y = "Gene Count") +
                            theme_bw() +
                            theme(plot.title = element_text(size = 10, 
                                                            face = "bold", 
                                                            hjust = 0.5))
jpeg("../Enrichment analysis in R/outputs/kegg_enriched_up_dotplot.jpeg",
width = 800, height = 800)
print(p)
dev.off()
}
dot_plot_kegg_enriched_up()
```



# Visualize a specific pathway

## visualize the top enriched KEGG pathway among up-regulated sig genes
```{r}
visualize_top_path_up <- function(){
  pathview(gene.data = up_entrez_ids, 
           pathway.id = "dme04512", 
           species = "dme",
           kegg.dir = "../Enrichment analysis in R/outputs/")
}
visualize_top_path_up()
```

## Browse the top enriched pathway among up-regulated sig genes
```{r}
browseKEGG(kegg_enrich_up, 'dme04512')
```



## Downregulated Dmel_XXXX ids 
```{r}
dmel_down <- function(){
  filtered_down_sig_genes <- gene_info_result[gene_info_result$ensembl_gene_id %in% rownames(downregulated_sig_genes), ]
  
  dmel_down_identifiers <- paste("Dmel", filtered_down_sig_genes$external_gene_name,sep = "_")
  return(dmel_down_identifiers)
}
dmel_down_genes <- dmel_down()
dmel_down_genes
```

## KEGG pathway enrichment analysis among down-regulated significant genes
```{r message=FALSE}
kegg_enrichment_down <- function(){
  kegg_down <- enrichKEGG(gene = dmel_down_genes,
                          organism = "dme", # Drosophila melanogaster
                          pvalueCutoff = 0.05,
                          pAdjustMethod = "BH")
  if (is.null(kegg_down) || nrow(kegg_down) == 0) {
      print("No enriched pathways found")
    }
  return(kegg_down)
}
kegg_enrich_down <- kegg_enrichment_down()
```

## Reactome pathway enrichment analysis among down-regulated significant genes
```{r}
reactome_down_genes <- function(){
  reactome_enrichment <- enrichPathway(gene = down_entrez_ids,
                                       organism = "fly", # Drosophila melanogaster
                                       pvalueCutoff = 0.05)
  
  return(reactome_enrichment)
}
reactome_enriched_path_down <- reactome_down_genes()
head(reactome_enriched_path_down)
```

## DataFrame of Reactome enriched pathways among down-regulated significant genes
```{r}
dataframe_reactome_down <- function(){
  df_reactome_down_genes <- as.data.frame(reactome_enriched_path_down)
  return(df_reactome_down_genes)
}
df_reactome_down <- dataframe_reactome_down()
head(df_reactome_down)
```


# Visualize Reactome Pathway Enrichment Results

## Barplot for Reactome enriched pathways among down-regulated significant gene 
```{r}
bar_plot_reactome_enriched_down<- function(){
jama_colors <- pal_jama("default")(7)
p <- ggplot(df_reactome_down, aes(x = reorder(Description, - Count) ,
                                  y = Count, fill = p.adjust)) +
                                  geom_bar(stat = "identity", width = 0.8) +
                                  coord_flip() +
                                  scale_fill_gradientn(name = "p.adjust", 
                                                       colors = jama_colors) + labs(title = "Bar plot of Enriched Reactome pathways among down-regulated Genes",
     x = "Enriched pathway",
     y = "Gene Count") +
                                  theme_bw() +
                                  theme(plot.title = element_text(size = 10, 
                                                                  face = "bold",                                                                     hjust = 0.5))
jpeg("../Enrichment analysis in R/outputs/reactome_enriched_down_barplot.jpeg",
width = 700, height = 800)
print(p)
dev.off()
}
bar_plot_reactome_enriched_down()
```

## Dotplot for Reactome enriched pathways among downregulated significant gene 
```{r}
dot_plot_reactome_enriched_down<- function(){
jama_colors <- pal_jama("default")(7)
p <- ggplot(df_reactome_down, aes(x = reorder(Description, -Count),
                                  y = Count, size = Count,
                                  color = p.adjust)) +
                                  geom_point(alpha = 0.6) +
                                  coord_flip() +
                                  scale_size_continuous(range = c(3, 8), 
                                                        name = "Gene Count") +
                                  scale_color_gradientn(name = "p.adjust", 
                                                        colors = jama_colors) +
labs(title = "Dot plot of enriched Reactome pathways among down-regulated genes",
     x = "GO Term",
     y = "Gene Count") +
                                  theme_bw() +
                                  theme(plot.title = element_text(size = 10, 
                                                                  face = "bold",                                                                     hjust = 0.5))
jpeg("../Enrichment analysis in R/outputs/reactome_enriched_down_dotplot.jpeg",
width = 700, height = 800)
print(p)
dev.off()
}
dot_plot_reactome_enriched_down()
```

## visualize the top enriched Reactome pathway for down-regulated sig genes
## Take the generated URL and browse it 
```{r}
visualize_reactome_path_up <- function(){
  # Convert the Reactome ID to a URL for visualization
  reactome_url <- paste0("https://reactome.org/PathwayBrowser/#/", "R-DME-8964038")
  # Print the URL for manual review
  print(reactome_url)
}
visualize_reactome_path_up()
```











